An Open Platform for Managing Short Programming Exercises

### 2013-06-18 paper/paper.tex
restore author info
comment out sec 5, fix citation

% \alignauthor
% Andrei Papancea\\
% Jaime Spacco\\
%        \affaddr{Department of Computer Science}\\
%        \affaddr{Knox College}\\
%        \affaddr{Galesburg, IL 61401}\\
%        \email{\{apapance,jspacco\}@knox.edu}
% % 2nd. author
% \alignauthor
% David Hovemeyer\\
%        \affaddr{Kinsley Engineering Center}\\
%        \affaddr{York College of Pennsylvania}\\
%        \affaddr{York, PA 17403}\\
%        \email{dhovemey@ycp.edu}

\author{
\alignauthor
Author Name\\
Author Name\\
       \affaddr{Department of Computer Science}\\
       \affaddr{Small College \#1}\\
       \affaddr{City, State zipcode}\\
       \email{\{author1,author2\}@school.edu}
% 2nd. author
\alignauthor
Author Name\\
       \affaddr{Computer Science Department}\\
       \affaddr{Small College \#2}\\
       \affaddr{City, State zipcode}\\
       \email{author@school.edu}
}

@@

Recent years have seen an explosion of online services for learning
to program:  Udacity \cite{udacity}, CodeAcademy \cite{codeacademy},
ProgZoo \cite{progzoo}, 
Codingbat \cite{codingbat}, Coursera \cite{coursera}, Turingscraft
\cite{turingscraft}, Problets \cite{Kumar::2005:GPA:1163405.1163408}, CS Circles \cite{Pritchard:2013:CCI:2445196.2445370} and
several others.  These systems are very powerful, and show great promise to
improve introductory computer science.  Given all of these
very good systems, one may ask why we thought it a good career move
to build yet another system.  There are three main reasons:

@@

\begin{figure*}[ht]
\centering
\includegraphics[width=180mm]{images/screenshot.png}
\caption{Screenshot of the main view in CloudCoder.}
\label{screenshot}
\end{figure*}

@@

\begin{table*}
\centering
\begin{tabular}{| l | l | l | l |}
\hline
Institution & Course & Language & Enrollment\\
\hline
Canisius College & CS-1 & C/C++ & 30\\
Small College \#2 & CS-1 & C/C++ & 120\\
Small College \#1 & CS-1 & Java & 23\\
University of Auckland & Programming for Engineers & C & 30\\
\hline
\end{tabular}
\caption{CloudCoder pilot studies.}
\label{tab:courses}
\end{table*}

@@

\section{Future Work: Towards an Intelligent Tutoring System}

One of the long-term goals for CloudCoder is to use the data to
develop an Intelligent Tutoring System for programming assignments.
A web-based programming ITS could be used to support high school
AP courses in school districts that may lack computer science
teachers, or to help improve introductory CS courses by reducing
student frustration.

To develop an ITS, we need a large supply of exercises; we hope
CloudCoder's permissible licensing will help on this front.  We also
need to provide students with ``hints'' when they are stuck.  To
provide hints, we can either write the hints ourselves, which is
labor-intensive, or generate them automatically by mining them out of
the dataset of prior submissions.  Obviously this second approach is
a much preferred method of creating hints.  However, identifying hints
automatically means that we need to match a student's current program
state with the most similar program state we've seen previously, 
figure out what successful next step the previous student took, and
then summarize this information into a hint.  This is a very
challenging problem.

There are currently two promising lines of research
trying to address the issue of identifying the most similar previous submissions.  Rivers and Koedinger \cite{rivers-its2012} have developed a
system that computes a {\em canonicalized} version of each submission
by compiling to an Abstract Syntax Tree (AST) and then applying a
series of compiler transformations.  This work is ongoing, but
preliminary results are encouraging.
On the other hand, Jin et. al. \cite{Jin:2012:PRA:2345840.2345889, jin-kdd2011} perform
an inter-procedural data flow analysis that tracks the values stored
in different variables in an effort to identify variables that are
used in the same way in each submission.  They also show promising preliminary
results for their approach.

\subsection{Key Features:  Mining Hints from the AST}

Our approach, which is very preliminary, ongoing research,  is also based on an analysis of
the AST.
Our hypothesis is that many novice-level exercises
will have solutions that require certain structural
{\em key features} in the code that we can detect automatically by
comparing the AST of the student's code with the ASTs of known correct
solutions.

For example, Figures~\ref{fig:counta} and \ref{fig:countab} show two
CloudCoder exercises that were assigned during the Winter 2013 term in
CS-1 at 
Small College \#1
%Knox College.

\begin{figure}[h]
{\tt Write a static method called countA that takes a String as a
parameter and returns the number of times that 'a' occurs in the String. }
\caption{CountA:  An example CloudCoder exercise.}
\label{fig:counta}
\end{figure}

\begin{figure}[h]
{\tt Write a static method called countAB that takes a String as a
parameter and returns the number of times that ab occurs (i.e. an 'a'
immediately followed by 'b'). }
\caption{CountAB: Another example CloudCoder exercise.}
\label{fig:countab}
\end{figure}

First let us consider the {\tt CountA} exercise from
Figure~\ref{fig:counta}.  As is typical in computer science, students have a lot of flexibility
in how they solve this problem:  with a for loop or with a while loop, using the {\tt charAt()}
method or the {\tt substring()} method or other String methods, and so on.  However, in
practice, exercises are assigned in a context with other
exercises, in-class examples, and course topics, which tends to limit
the scope of possible student solutions, and to push students towards
certain solutions and away from others.  For example, no student solved this
exercise with a while loop, and all but one used {\tt charAt()}
rather than other String operations.  In fact, most correct
solutions looked something like Figure~\ref{fig:counta-solution}.

\begin{figure}[h]
\begin{lstlisting}
public static int countAB(String str) {
  int count=0;
  for (int i=0; i<str.length();i++) {
    if (str.charAt(i)=='a') {
      count++;
    }
  }
  return count;
}
\end{lstlisting}
\caption{Typical solution for Exercise CountA depicted in Figure~\ref{fig:counta}.}
\label{fig:counta-solution}
\end{figure}

\begin{figure}
\includegraphics[width=90mm]{images/countA.pdf}
\caption{Slightly simplified AST of the code in Figure~\ref{fig:counta-solution}}
\label{fig:counta-ast}
\end{figure}

Compiling the method from Figure~\ref{fig:counta-solution} to an
Abstract Syntax Tree (AST) produces a tree that looks something like
Figure~\ref{fig:counta-ast} (the AST has been simplified slightly by
collapsing nodes that are unimportant for this example).  We can identify
several features of the AST that must be present to solve the
exercise, such as a for loop, an if-statement inside the loop, and
the use of a String method, typically {\tt charAt()}, to examine
characters inside the loop.  We can also ensure that the components of
the for loop are correct, i.e. that the loop counter starts at 0, that
the loop goes up to the length of the String, and so on.  Assuming
that student code compiles, all of these necessary features can be
extracted from the AST.

Now let's consider the {\tt CountAB} exercise depicted in
Figure~\ref{fig:countab}.  Again, there were many correct solutions;
however, many correct solutions looked something like
Figure~\ref{fig:countab-solution}.

\begin{figure}[h]
\begin{lstlisting}
public static int countAB(String str) {
  int count=0;
  for (int i=0; i<str.length()-1;i++) {
    if (str.charAt(i)=='a' && 
        str.charAt(i+1)=='b') {
      count++;
    }
  }
  return count;
}
\end{lstlisting}
\caption{Typical solution for Exercise countAB depicted in Figure~\ref{fig:countab}.}
\label{fig:countab-solution}
\end{figure}

Compiling the method in Figure~\ref{fig:countab-solution} into an
abstract syntax tree (AST) produces a tree that looks like
Figure~\ref{fig:countab-ast} (this result has been slightly simplified
from the full output from the Eclipse JDT compiler).

\begin{figure}[h]
\includegraphics[width=90mm]{images/forloop-highlight.png}
\caption{Slightly simplified AST of the code in Figure~\ref{fig:countab-solution}}
\label{fig:countab-ast}
\end{figure}

Note the two circled sub-trees, or {\em features}, in the AST depicted
in Figure~\ref{fig:countab-ast}.  Feature \#1, the {\em
  length-1} feature, shows the loop iterating
until the length of the String minus one.  The for loop requires a
stop condition that is different
than the loop stop condition used to solve the {\tt CountA} exercise from
Figure~\ref{fig:counta}.  Crucially, once we know students are
using a for loop to solve this exercise (which we can also
extract from the AST), {\em we know exactly where in the AST to
  look for the length-1 feature}.  If a submission has all the
features except the {\em length-1} feature, we have a pretty good idea
that the student hasn't figured out to change the stop condition of the
for loop.  We could use the lack of features present in correct
solutions to give us an idea of what things the student is missing.

\subsection{Limits of Key Features Approach}

Our approach of identifying ``key features'' has a number of
weaknesses and limitations.  First, like other ITS hint-systems, it
requires a reliable system for identifying the nearest previous successful
submission to a student's current program state.  Work in this area is
ongoing \cite{rivers-its2012,Jin:2012:PRA:2345840.2345889}.

Second, features are based purely on the structure of the
code, and are treated independently of one another, which is a
simplification that throws out a lot of relevant contextual information.  For
example, a student submission may have two for loops, one of which has
an if statement nested inside it; our approach cannot determine which
loop has the if statement inside it.

Third, we do not perform any data-flow analysis (at least not yet), so
we cannot identify if particular values are stored in variables.

Next, we require compilable code, which can be a problem for the many
students who struggle to get their code to compile
\cite{Denny:2012:SEE:2325296.2325318}.

Finally, while we focused mainly on exercises that required for loops
to solve, we also assigned a number of exercises that
use if statements and no loops.  For these exercises, we can figure
out when students don't have 
enough structure, such as when they don't have nested if statements
for an exercise where most people who solves the exercise used nested if statements.  However,
we cannot yet easily determine when students have the wrong boundary
conditions in an if statement.

Much more work is needed in this area.  We hope that 
CloudCoder can help out by providing a rich data set for post-hoc
studies, as well as a platform for testing out a hint-system.

>>>>

\author{
\alignauthor
Andrei Papancea\\
Jaime Spacco\\
       \affaddr{Department of Computer Science}\\
       \affaddr{Knox College}\\
       \affaddr{Galesburg, IL 61401}\\
       \email{\{apapance,jspacco\}@knox.edu}
% 2nd. author
\alignauthor
David Hovemeyer\\
       \affaddr{Kinsley Engineering Center}\\
       \affaddr{York College of Pennsylvania}\\
       \affaddr{York, PA 17403}\\
       \email{dhovemey@ycp.edu}
}

%\author{
%\alignauthor
%Author Name\\
%Author Name\\
%       \affaddr{Department of Computer Science}\\
%       \affaddr{Small College \#1}\\
%       \affaddr{City, State zipcode}\\
%       \email{\{author1,author2\}@school.edu}
%% 2nd. author
%\alignauthor
%Author Name\\
%       \affaddr{Computer Science Department}\\
%       \affaddr{Small College \#2}\\
%       \affaddr{City, State zipcode}\\
%       \email{author@school.edu}
%}

@@

Recent years have seen an explosion of online services for learning
to program:  Udacity \cite{udacity}, CodeAcademy \cite{codeacademy},
ProgZoo \cite{progzoo}, 
Codingbat \cite{codingbat}, Coursera \cite{coursera}, Turingscraft
\cite{turingscraft}, Problets \cite{Kumar:2005:GPA:1163405.1163408}, CS Circles \cite{Pritchard:2013:CCI:2445196.2445370} and
several others.  These systems are very powerful, and show great promise to
improve introductory computer science.  Given all of these
very good systems, one may ask why we thought it a good career move
to build yet another system.  There are three main reasons:

@@

\begin{figure*}[ht]
\centering
\includegraphics[width=180mm]{images/screenshot}
\caption{Screenshot of the main view in CloudCoder.}
\label{screenshot}
\end{figure*}

@@

\begin{table*}
\centering
\begin{tabular}{| l | l | l | l |}
\hline
Institution & Course & Language & Enrollment\\
\hline
Canisius College & CS-1 & C/C++ & 30\\
York College & CS-1 & C/C++ & 120\\
Knox College \#1 & CS-1 & Java & 23\\
University of Auckland & Programming for Engineers & C & 30\\
\hline
\end{tabular}
\caption{CloudCoder pilot studies.}
\label{tab:courses}
\end{table*}

@@

%
%\section{Future Work: Towards an Intelligent Tutoring System}
%
%One of the long-term goals for CloudCoder is to use the data to
%develop an Intelligent Tutoring System for programming assignments.
%A web-based programming ITS could be used to support high school
%AP courses in school districts that may lack computer science
%teachers, or to help improve introductory CS courses by reducing
%student frustration.
%
%To develop an ITS, we need a large supply of exercises; we hope
%CloudCoder's permissible licensing will help on this front.  We also
%need to provide students with ``hints'' when they are stuck.  To
%provide hints, we can either write the hints ourselves, which is
%labor-intensive, or generate them automatically by mining them out of
%the dataset of prior submissions.  Obviously this second approach is
%a much preferred method of creating hints.  However, identifying hints
%automatically means that we need to match a student's current program
%state with the most similar program state we've seen previously, 
%figure out what successful next step the previous student took, and
%then summarize this information into a hint.  This is a very
%challenging problem.
%
%There are currently two promising lines of research
%trying to address the issue of identifying the most similar previous submissions.  Rivers and Koedinger \cite{rivers-its2012} have developed a
%system that computes a {\em canonicalized} version of each submission
%by compiling to an Abstract Syntax Tree (AST) and then applying a
%series of compiler transformations.  This work is ongoing, but
%preliminary results are encouraging.
%On the other hand, Jin et. al. \cite{Jin:2012:PRA:2345840.2345889, jin-kdd2011} perform
%an inter-procedural data flow analysis that tracks the values stored
%in different variables in an effort to identify variables that are
%used in the same way in each submission.  They also show promising preliminary
%results for their approach.
%
%\subsection{Key Features:  Mining Hints from the AST}
%
%Our approach, which is very preliminary, ongoing research,  is also based on an analysis of
%the AST.
%Our hypothesis is that many novice-level exercises
%will have solutions that require certain structural
%{\em key features} in the code that we can detect automatically by
%comparing the AST of the student's code with the ASTs of known correct
%solutions.
%
%For example, Figures~\ref{fig:counta} and \ref{fig:countab} show two
%CloudCoder exercises that were assigned during the Winter 2013 term in
%CS-1 at 
%Small College \#1
%%Knox College.
%
%\begin{figure}[h]
%{\tt Write a static method called countA that takes a String as a
%parameter and returns the number of times that 'a' occurs in the String. }
%\caption{CountA:  An example CloudCoder exercise.}
%\label{fig:counta}
%\end{figure}
%
%
%\begin{figure}[h]
%{\tt Write a static method called countAB that takes a String as a
%parameter and returns the number of times that ab occurs (i.e. an 'a'
%immediately followed by 'b'). }
%\caption{CountAB: Another example CloudCoder exercise.}
%\label{fig:countab}
%\end{figure}
%
%First let us consider the {\tt CountA} exercise from
%Figure~\ref{fig:counta}.  As is typical in computer science, students have a lot of flexibility
%in how they solve this problem:  with a for loop or with a while loop, using the {\tt charAt()}
%method or the {\tt substring()} method or other String methods, and so on.  However, in
%practice, exercises are assigned in a context with other
%exercises, in-class examples, and course topics, which tends to limit
%the scope of possible student solutions, and to push students towards
%certain solutions and away from others.  For example, no student solved this
%exercise with a while loop, and all but one used {\tt charAt()}
%rather than other String operations.  In fact, most correct
%solutions looked something like Figure~\ref{fig:counta-solution}.
%
%\begin{figure}[h]
%\begin{lstlisting}
%public static int countAB(String str) {
%  int count=0;
%  for (int i=0; i<str.length();i++) {
%    if (str.charAt(i)=='a') {
%      count++;
%    }
%  }
%  return count;
%}
%\end{lstlisting}
%\caption{Typical solution for Exercise CountA depicted in Figure~\ref{fig:counta}.}
%\label{fig:counta-solution}
%\end{figure}
%
%
%\begin{figure}
%\includegraphics[width=90mm]{images/countA.pdf}
%\caption{Slightly simplified AST of the code in Figure~\ref{fig:counta-solution}}
%\label{fig:counta-ast}
%\end{figure}
%
%Compiling the method from Figure~\ref{fig:counta-solution} to an
%Abstract Syntax Tree (AST) produces a tree that looks something like
%Figure~\ref{fig:counta-ast} (the AST has been simplified slightly by
%collapsing nodes that are unimportant for this example).  We can identify
%several features of the AST that must be present to solve the
%exercise, such as a for loop, an if-statement inside the loop, and
%the use of a String method, typically {\tt charAt()}, to examine
%characters inside the loop.  We can also ensure that the components of
%the for loop are correct, i.e. that the loop counter starts at 0, that
%the loop goes up to the length of the String, and so on.  Assuming
%that student code compiles, all of these necessary features can be
%extracted from the AST. 
%
%Now let's consider the {\tt CountAB} exercise depicted in
%Figure~\ref{fig:countab}.  Again, there were many correct solutions;
%however, many correct solutions looked something like
%Figure~\ref{fig:countab-solution}.  
%
%\begin{figure}[h]
%\begin{lstlisting}
%public static int countAB(String str) {
%  int count=0;
%  for (int i=0; i<str.length()-1;i++) {
%    if (str.charAt(i)=='a' && 
%        str.charAt(i+1)=='b') {
%      count++;
%    }
%  }
%  return count;
%}
%\end{lstlisting}
%\caption{Typical solution for Exercise countAB depicted in Figure~\ref{fig:countab}.}
%\label{fig:countab-solution}
%\end{figure}
%
%Compiling the method in Figure~\ref{fig:countab-solution} into an
%abstract syntax tree (AST) produces a tree that looks like
%Figure~\ref{fig:countab-ast} (this result has been slightly simplified
%from the full output from the Eclipse JDT compiler).
%
%\begin{figure}[h]
%\includegraphics[width=90mm]{images/forloop-highlight}
%\caption{Slightly simplified AST of the code in Figure~\ref{fig:countab-solution}}
%\label{fig:countab-ast}
%\end{figure}
%
%Note the two circled sub-trees, or {\em features}, in the AST depicted
%in Figure~\ref{fig:countab-ast}.  Feature \#1, the {\em
%  length-1} feature, shows the loop iterating
%until the length of the String minus one.  The for loop requires a
%stop condition that is different
%than the loop stop condition used to solve the {\tt CountA} exercise from
%Figure~\ref{fig:counta}.  Crucially, once we know students are
%using a for loop to solve this exercise (which we can also
%extract from the AST), {\em we know exactly where in the AST to
%  look for the length-1 feature}.  If a submission has all the
%features except the {\em length-1} feature, we have a pretty good idea
%that the student hasn't figured out to change the stop condition of the
%for loop.  We could use the lack of features present in correct
%solutions to give us an idea of what things the student is missing.
%
%
%\subsection{Limits of Key Features Approach}
%
%Our approach of identifying ``key features'' has a number of
%weaknesses and limitations.  First, like other ITS hint-systems, it
%requires a reliable system for identifying the nearest previous successful
%submission to a student's current program state.  Work in this area is
%ongoing \cite{rivers-its2012,Jin:2012:PRA:2345840.2345889}.
%
%Second, features are based purely on the structure of the
%code, and are treated independently of one another, which is a
%simplification that throws out a lot of relevant contextual information.  For
%example, a student submission may have two for loops, one of which has
%an if statement nested inside it; our approach cannot determine which
%loop has the if statement inside it.  
%
%Third, we do not perform any data-flow analysis (at least not yet), so
%we cannot identify if particular values are stored in variables.
%
%Next, we require compilable code, which can be a problem for the many
%students who struggle to get their code to compile
%\cite{Denny:2012:SEE:2325296.2325318}.
%
%Finally, while we focused mainly on exercises that required for loops
%to solve, we also assigned a number of exercises that
%use if statements and no loops.  For these exercises, we can figure
%out when students don't have 
%enough structure, such as when they don't have nested if statements
%for an exercise where most people who solves the exercise used nested if statements.  However,
%we cannot yet easily determine when students have the wrong boundary
%conditions in an if statement.
%
%Much more work is needed in this area.  We hope that 
%CloudCoder can help out by providing a rich data set for post-hoc
%studies, as well as a platform for testing out a hint-system.
